---
title: "Benchmark: binnr versus standard R learners"
author:
  - name: David Selby
    affiliation: Deutsches Forschungszentrum für Künstliche Intelligenz (DFKI)
    email: david_antony.selby@dfki.de
  - name: Daria Pavliuk
    affiliation: DFKI
  - name: Gerrit Grossmann
    affiliation: DFKI
  - name: Sebastian Vollmer
    affiliation: DFKI
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Benchmark: binnr versus standard R learners}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7.5, fig.height = 4,
                      dpi = 96, echo = FALSE)
```

This vignette summarises a small cross-validated comparison between binnr
models and learners commonly used for omics prediction in R. The results are
**precomputed** by `inst/scripts/benchmark.R` (about three hours on a 2-core
laptop), because refitting hundreds of models is too slow for a vignette. Run it
yourself with

```sh
Rscript inst/scripts/benchmark.R all 5 5
```

# Setup

* **Data**: the TCGA BRCA and PRAD example cohorts from `tcga_cohort()`
  (1,000 Reactome genes per cohort).
* **Tasks**:
  - *PAM50 subtype* (BRCA): 4-class classification from RNA, copy number and
    mutation (n = 484), scored by accuracy and balanced accuracy;
  - *Overall survival* (BRCA): Cox model from omics plus age and stage (n = 928,
    about 130 deaths), scored by Harrell's C-index;
  - *Gleason grade* (PRAD): high (8–10) versus low (6–7) grade from RNA and copy
    number (n = 491, 41% high), the task used to showcase P-NET, scored by AUROC.
* **Evaluation**: 5-fold cross-validation (stratified by class or by event
  status), repeated 5 times, with identical folds for every method (25 splits).
* **Tuning**: every learner chooses its hyperparameters *inside* the training
  fold, with a matched budget. binnr models select the graph depth (5 levels,
  as P-NET truncates Reactome, or the full hierarchy), and the weight decay
  (BINN) or state dimension (GNN), by the early-stopping validation loss via
  `bn_tune()`; `cv.glmnet` selects its penalty and `xgb.cv` the number of
  boosting rounds by inner 5-fold CV; ranger runs at its defaults. The graph
  variants below reuse the configuration selected for the Reactome BINN.
* **Uncertainty**: comparison intervals from *quasi-variances* (Firth & de
  Menezes, 2004; `qvcalc`) of the method effects in a split-blocked linear
  model `score ~ method + split`. Two methods differ at roughly the 5% level
  when their intervals (mean ± 1.39 quasi-standard errors) do not overlap.
  Because the 25 splits are not independent blocks (the same samples recur and
  training sets overlap), the contrast covariance is inflated by the
  Nadeau & Bengio (2003) factor $1 + J/(k-1)$ before computing the
  quasi-variances; this roughly triples the interval width and is the
  conservative choice.
* **Methods**:
  - `Lasso (glmnet)`: L1-penalised multinomial or Cox regression, with the
    penalty chosen by inner 5-fold CV. In the survival task, age and stage are
    unpenalised.
  - `Random forest (ranger)`: 500 trees; a survival forest for the Cox task,
    with risk scored as the summed cumulative hazard.
  - `Boosting (xgboost)`: 300 rounds, learning rate 0.05 (0.03 for survival),
    depth 3 (2), 30% column subsampling.
  - `Cox, clinical only`: `survival::coxph(Surv ~ age + stage)`.
  - `BINN (binnr)`: `bn_arch("binn")`, the classic BINN architecture
    (scalar pathway nodes, one weight per edge, readout from the root pathways),
    trained with binnr's topological schedule on the unpadded Reactome DAG.
  - `BINN, stratified (padded)`: the same architecture on the padded graph
    from `bn_pad()`, i.e. P-NET's stratified construction with copy nodes.
  - `GNN (binnr)`: `bn_arch("gnn")`, with 4-dimensional pathway embeddings and
    GCN-style shared weights.
  - `BINN, rewired graph`: a BINN on a degree-preserving randomisation of the
    Reactome graph (`bn_rewire("degree")`), with the same sparsity and degrees
    but random gene–pathway assignments. A new null graph is drawn for every
    split.
  - `BINN, random graph`: a BINN on an Erdős–Rényi graph with the same number
    of gene–pathway edges (`bn_rewire("bernoulli")`), redrawn for every split.
  - `BINN, 50% of annotations removed`: a BINN on an *incomplete* knowledge
    graph (`bn_drop_edges(prop = 0.5)`), redrawn for every split.

  All binnr models use the default `bn_control()`: AdamW, early stopping
  on a 20% validation split, and no tuning.

```{r load}
bench_file <- system.file("extdata", "benchmark.rds", package = "binnr")
have <- nzchar(bench_file) && requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(eval = have)
```

```{r data, eval = have}
bench <- readRDS(bench_file)
info <- attr(bench, "info")
order_methods <- c("Cox, clinical only", "Lasso (glmnet)", "Random forest (ranger)",
                   "Boosting (xgboost)", "BINN (binnr)", "BINN, stratified (padded)",
                   "GNN (binnr)", "BINN, rewired graph", "BINN, random graph",
                   "BINN, 50% of annotations removed")
bench$task <- factor(bench$task, levels = c("PAM50 subtype", "Overall survival", "Gleason grade"))
bench$method <- factor(bench$method, levels = rev(intersect(order_methods, bench$method)))
family_of <- function(m) ifelse(grepl("binnr|BINN|GNN", m), "binnr", "baseline")
```

# Results

```{r helpers, eval = have}
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
k <- info$folds
# Corrected resampled t interval (Nadeau & Bengio, 2003), for paired differences
corrected_t <- function(x, k) {
  J <- length(x)
  se <- sqrt((1 / J + 1 / (k - 1)) * stats::var(x))
  q <- stats::qt(0.975, J - 1)
  tibble(mean = mean(x), lower = mean(x) - q * se, upper = mean(x) + q * se,
         p = 2 * stats::pt(-abs(mean(x) / se), J - 1))
}
# Quasi-variance comparison intervals (Firth & de Menezes, 2004) from the
# split-blocked model, with the Nadeau-Bengio inflation of the covariance
qv_summary <- function(d, k, inflate = TRUE) {
  d <- d |> mutate(split = interaction(rep, fold, drop = TRUE), method = droplevels(method))
  fit <- lm(value ~ method + split, data = d)
  m <- nlevels(d$method); J <- nlevels(d$split)
  idx <- grep("^method", names(coef(fit)))
  Vfull <- matrix(0, m, m); Vfull[-1, -1] <- vcov(fit)[idx, idx]
  if (inflate) Vfull <- Vfull * (1 + J / (k - 1))
  qv <- qvcalc::qvcalc(Vfull, estimates = c(0, coef(fit)[idx]), labels = levels(d$method))$qvframe
  tibble(method = factor(levels(d$method), levels(d$method)),
         mean = as.numeric(tapply(d$value, d$method, mean)[levels(d$method)]),
         qse = sqrt(qv$quasiVar)) |>
    mutate(lower = mean - 1.39 * qse, upper = mean + 1.39 * qse)
}
metric_labels <- c(accuracy = "Accuracy", balanced_accuracy = "Balanced accuracy",
                   c_index = "C-index", auroc = "AUROC")
# Headline metric per task, for the figures
headline <- bench[(bench$task == "PAM50 subtype" & bench$metric == "accuracy") |
                  (bench$task == "Overall survival" & bench$metric == "c_index") |
                  (bench$task == "Gleason grade" & bench$metric == "auroc"), ]
```

```{r table, eval = have}
summ <- bench |>
  group_by(task, metric) |>
  group_modify(~ qv_summary(.x, k)) |>
  ungroup() |>
  left_join(bench |> distinct(task, method, rep, fold, seconds) |>
              group_by(task, method) |> summarise(seconds = median(seconds), .groups = "drop"),
            by = c("task", "method")) |>
  arrange(task, metric, desc(mean))
summ |>
  mutate(`comparison interval` = sprintf("[%.3f, %.3f]", lower, upper),
         metric = metric_labels[metric]) |>
  select(task, metric, method, mean, QSE = qse, `comparison interval`, `median s` = seconds) |>
  knitr::kable(digits = 3, caption = sprintf(
    "Mean over %d x %d-fold cross-validation with quasi-standard errors and comparison intervals (mean +/- 1.39 QSE).",
    info$repeats, info$folds))
```

```{r plot, eval = have, fig.height = 5.5}
summ |>
  semi_join(distinct(headline, task, metric), by = c("task", "metric")) |>
  mutate(metric = metric_labels[metric]) |>
  ggplot(aes(mean, method, colour = family_of(method))) +
  geom_pointrange(aes(xmin = lower, xmax = upper)) +
  facet_wrap(~ task + metric, scales = "free_x") +
  scale_colour_manual(values = c(baseline = "#7A7A7A", binnr = "#3B6EA8")) +
  labs(x = "Mean with quasi-variance comparison interval", y = NULL, colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Paired comparisons with the Reactome BINN

Differences are computed within each split, which removes the variability
shared by all methods (easy and hard folds), with corrected resampled *t*
intervals (Nadeau & Bengio, 2003). Positive values mean the method beats the
Reactome BINN.

```{r paired, eval = have}
ref <- "BINN (binnr)"
diffs <- bench |>
  filter(method != ref) |>
  inner_join(bench |> filter(method == ref) |> select(task, metric, rep, fold, ref = value),
             by = c("task", "metric", "rep", "fold")) |>
  mutate(diff = value - ref)
paired <- diffs |>
  group_by(task, metric, method) |>
  reframe(corrected_t(diff, k)) |>
  arrange(task, metric, desc(mean))
paired |>
  mutate(`95% CI` = sprintf("[%+.3f, %+.3f]", lower, upper),
         metric = metric_labels[metric]) |>
  select(task, metric, method, difference = mean, `95% CI`, p) |>
  knitr::kable(digits = 3, caption = "Difference from the Reactome BINN (corrected resampled t).")
```

```{r paired-plot, eval = have, fig.height = 5}
paired |>
  semi_join(distinct(headline, task, metric), by = c("task", "metric")) |>
  mutate(metric = metric_labels[metric]) |>
  ggplot(aes(mean, method, colour = family_of(method))) +
  geom_vline(xintercept = 0, colour = "grey40") +
  geom_pointrange(aes(xmin = lower, xmax = upper)) +
  facet_wrap(~ task + metric, scales = "free_x") +
  scale_colour_manual(values = c(baseline = "#7A7A7A", binnr = "#3B6EA8")) +
  labs(x = "Difference from Reactome BINN (95% interval)", y = NULL, colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Which configurations were selected?

```{r selected, eval = have && !is.null(attr(bench, "selected"))}
attr(bench, "selected") |>
  mutate(depth = ifelse(is.finite(max_level), "5 levels", "full")) |>
  count(task, method, depth, hidden_dim, weight_decay) |>
  arrange(task, method, desc(n)) |>
  knitr::kable(caption = "Configurations chosen by validation loss, by task and split count.")
```

# Interpretation

```{r numbers, eval = have}
est <- function(tsk, met, mth, tab = summ) {
  r <- tab[tab$task == tsk & tab$metric == met & tab$method == mth, ]
  sprintf("%.3f [%.3f, %.3f]", r$mean, r$lower, r$upper)
}
dif <- function(tsk, met, mth) est(tsk, met, mth, paired)
```

* **No free lunch from pathways.** With matched within-fold tuning the Reactome BINN
  reaches a PAM50 accuracy of `r est("PAM50 subtype", "accuracy", "BINN (binnr)")`,
  a survival C-index of `r est("Overall survival", "c_index", "BINN (binnr)")`
  and a Gleason AUROC of `r est("Gleason grade", "auroc", "BINN (binnr)")`.
  The lasso differs from it by
  `r dif("PAM50 subtype", "accuracy", "Lasso (glmnet)")` in accuracy,
  `r dif("Overall survival", "c_index", "Lasso (glmnet)")` in C-index and
  `r dif("Gleason grade", "auroc", "Lasso (glmnet)")` in AUROC (paired
  differences with 95% intervals; positive favours the lasso).
* **Stratified versus topological.** P-NET's padded construction differs from
  the unpadded DAG by
  `r dif("PAM50 subtype", "accuracy", "BINN, stratified (padded)")` (accuracy),
  `r dif("Overall survival", "c_index", "BINN, stratified (padded)")` (C-index) and
  `r dif("Gleason grade", "auroc", "BINN, stratified (padded)")` (AUROC). The
  differences are small and not consistent in sign, so the thousands of copy
  nodes buy no reliable accuracy. They are not free of consequences either: a
  copy node with its own weight and bias is not a skip connection, so the
  padded graph is a different (larger) model, which may explain the survival
  result.
* **Biology or sparsity?** A BINN on a freshly rewired null graph in every split
  differs from the Reactome BINN by
  `r dif("PAM50 subtype", "accuracy", "BINN, rewired graph")` (accuracy),
  `r dif("Overall survival", "c_index", "BINN, rewired graph")` (C-index) and
  `r dif("Gleason grade", "auroc", "BINN, rewired graph")` (AUROC); a fully
  random graph of the same density by
  `r dif("PAM50 subtype", "accuracy", "BINN, random graph")`,
  `r dif("Overall survival", "c_index", "BINN, random graph")` and
  `r dif("Gleason grade", "auroc", "BINN, random graph")`; and removing half
  of the annotations by
  `r dif("PAM50 subtype", "accuracy", "BINN, 50% of annotations removed")`,
  `r dif("Overall survival", "c_index", "BINN, 50% of annotations removed")` and
  `r dif("Gleason grade", "auroc", "BINN, 50% of annotations removed")`.
  The data give no evidence that the model uses the *specific* pathway
  knowledge. This echoes "sparsity is all you need", with one caveat: a null
  control is informative only if the model could exploit correct knowledge in
  the first place, which is not guaranteed (see `?bn_rewire`).
* **Omics add little to survival here.** Age and stage alone
  (C-index `r est("Overall survival", "c_index", "Cox, clinical only")`) are as
  good as any omics model.
* **Relaxing the architecture does not help by default.** The GNN preset is
  not better at n < 1,000, and it is slower. Its appeal is flexibility, and the
  ability to test hypotheses like the ones above, not accuracy out of the box.

These results come from two small datasets and a small tuning grid, and
should be read as a demonstration of the workflow, not as a definitive
comparison. The benchmark script can be adapted to other `curatedTCGAData`
cohorts or knowledge graphs.

```{r info, eval = have}
str(info)
```
