---
title: "Getting started with binnr"
author:
  - name: David Selby
    affiliation: Deutsches Forschungszentrum für Künstliche Intelligenz (DFKI)
    email: david_antony.selby@dfki.de
  - name: Daria Pavliuk
    affiliation: DFKI
  - name: Gerrit Grossmann
    affiliation: DFKI
  - name: Sebastian Vollmer
    affiliation: DFKI
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Getting started with binnr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  dpi = 96
)
run <- torch::torch_is_installed()
knitr::opts_chunk$set(eval = run)
```

# Introduction

*Biologically informed neural networks* (BINNs) are neural networks whose
wiring follows prior biological knowledge. In the most common design, omics features are
connected to the genes they measure, genes to the pathways they belong to, and
pathways to their parent pathways, e.g. in [Reactome](https://reactome.org). The
aim is a model that predicts well from high-dimensional omics data while its
hidden nodes remain interpretable as biological processes.

Existing BINN software is written in Python and implements the network as a
stack of *masked* dense layers. That hard-codes several assumptions:

* the hierarchy has a fixed number of layers, so shallow branches must be padded
  with "dummy" copy nodes;
* each pathway is represented by a single number;
* each edge has its own weight;
* other data (clinical variables, several omics layers) must be bolted on.

**binnr** instead treats the knowledge graph as *data* and fits a
message-passing graph neural network on it. The classic BINN is a special case,
the default in fact, and each of the assumptions above becomes an argument you
can change. Nothing is specific to Reactome, or to biology: any directed
acyclic graph from features to higher-level concepts can be supplied as an
edge list. The package works with standard R and Bioconductor objects (matrices, data
frames, `survival::Surv`, `MultiAssayExperiment`) and returns tidy output.

```{r load}
library(binnr)
library(survival)
```

# Example data

The examples use a multi-omics subset of TCGA breast cancer built by
`tcga_cohort("BRCA")` from `r BiocStyle::Biocpkg("curatedTCGAData")`. The
first call downloads the assays through ExperimentHub and caches a compact
processed copy (in `tcga_cache_dir()`), so the data are not shipped with the
package; later calls read the cache. The cohort has 960 primary tumours and
1,000 Reactome genes measured in three layers: RNA-seq expression, copy number
and somatic mutations. The clinical table includes age, stage, the PAM50
molecular subtype and overall survival.

```{r data-check, include = FALSE}
# The remaining chunks need the cached cohort (no download during R CMD check)
has_data <- tcga_cached("BRCA") && tcga_cached("PRAD")
knitr::opts_chunk$set(eval = run && has_data)
```

```{r data}
brca <- tcga_cohort("BRCA")
str(brca$omics, max.level = 1)
brca$clinical
```

# The knowledge graph

`pathway_graph()` builds a graph from any gene sets and an optional hierarchy.
Here we use the bundled Reactome snapshot. For the full, current release use
`reactome_graph()`; for MSigDB collections, pass the output of
`msigdbr::msigdbr()` directly.

```{r graph}
g <- pathway_graph(
  reactome_tcga$gene_sets,
  reactome_tcga$hierarchy,
  labels = reactome_tcga$pathways
)
g
summary(g)
```

A graph is just two tibbles, so it can be inspected and filtered with the usual
tools:

```{r graph-tables}
bn_nodes(g)
bn_edges(g)
```

## The cost of strata

Reactome is a *ragged* hierarchy: `r summary(g)$skip_edges` of its edges skip one
or more levels, and the root pathways sit at different depths. A feedforward
BINN updates the network one *stratum* (layer) at a time, so it must pad every
such path with copy nodes until the graph is strictly stratified. `bn_pad()`
performs that transformation, so its cost is easy to see:

```{r pad}
padded <- bn_pad(g)
rbind(original = summary(g), padded = summary(padded))
```

Padding adds `r summary(padded)$n_dummy` artificial nodes with no biological
meaning. binnr never needs them: its *topological* schedule updates each node
once, after its children, so a gene linked directly to a high-level pathway
simply sends its message there. (You can still fit the padded graph to
reproduce a stratified BINN exactly; the benchmark vignette does.)

# Classification: PAM50 subtypes

The formula interface takes the outcome on the left-hand side and clinical
covariates on the right (`~ 1` for none). `data` holds those variables, and
`omics` is a named list of matrices whose rows correspond to the rows of `data`.
Samples with missing values are dropped with a message, as in `lm()`.

```{r split}
set.seed(42)
test <- sample(nrow(brca$clinical), 300)
train <- setdiff(seq_len(nrow(brca$clinical)), test)
rows <- function(i) lapply(brca$omics, function(m) m[i, ])
```

```{r pam50-fit}
fit <- bn_fit(
  pam50 ~ 1,
  data = brca$clinical[train, ],
  omics = rows(train),
  graph = g,
  control = bn_control(seed = 1)
)
fit
```

With several omics layers, each feature becomes an input node (`TP53|rna`,
`TP53|cnv`, `TP53|mut`) feeding a gene node, as in P-NET. The default
architecture is a classic BINN:

```{r arch}
bn_arch()
```

`predict()` returns a tibble using the tidymodels naming conventions:

```{r pam50-predict}
test_ok <- test[!is.na(brca$clinical$pam50[test])]
pred <- predict(fit, rows(test_ok))
table(predicted = pred$.pred_class, observed = brca$clinical$pam50[test_ok])
mean(pred$.pred_class == brca$clinical$pam50[test_ok])
head(predict(fit, rows(test_ok), type = "prob"))
```

Training stops early when the loss on a held-out validation fraction (20% by
default) stops improving, and the best weights are restored:

```{r history, fig.height = 3.5}
ggplot2::autoplot(fit)
```

## Which pathways matter?

`node_importance()` scores every node (input features, genes and pathways) by
the mean absolute gradient × activation, a first-order attribution related to
the DeepLIFT and SHAP scores used by P-NET and BINN:

```{r importance}
imp <- node_importance(fit, rows(test_ok))
imp[imp$type == "pathway", c("label", "level", "importance")]
```

```{r importance-plot, fig.height = 4.5}
ggplot2::autoplot(imp, n = 12)
```

The top genes and features can be found the same way:

```{r importance-genes}
head(imp[imp$type == "input", c("node", "modality", "importance")], 10)
```

As the paper discusses, such explanations can vary between equally accurate
fits (the Rashomon effect). `bn_importance_stability()` refits the model with
several seeds and summarises how stable each pathway's rank is. Check this
before interpreting any single pathway.

## Pathway activity scores for downstream statistics

With scalar pathway nodes, each pathway has one activation per patient, which
can be read as a learned "pathway activity score". `node_activations()`
returns these as an ordinary samples × nodes matrix, so the usual tools for
wrangling, testing and plotting apply. For example, which pathways'
activities differ most between subtypes in the held-out patients?

```{r activations}
library(dplyr, warn.conflicts = FALSE)
library(tidyr)
act <- node_activations(fit, rows(test_ok), type = "pathway")
labels <- attr(act, "labels")
subtype <- brca$clinical$pam50[test_ok]

kw <- tibble(node = colnames(act), label = labels[colnames(act)]) |>
  mutate(
    p = apply(act, 2, function(a) {
      if (stats::sd(a) < 1e-8) NA_real_ else kruskal.test(a, subtype)$p.value
    })
  ) |>
  mutate(p_adj = p.adjust(p, method = "BH")) |>
  arrange(p)
kw
```

```{r activation-plot, fig.height = 4.5}
top <- head(kw$node, 4)
as_tibble(act[, top], rownames = "patient") |>
  mutate(subtype = subtype) |>
  pivot_longer(
    -c(patient, subtype),
    names_to = "node",
    values_to = "activity"
  ) |>
  mutate(pathway = labels[node]) |>
  ggplot2::ggplot(ggplot2::aes(subtype, activity, fill = subtype)) +
  ggplot2::geom_boxplot(outlier.size = 0.5, show.legend = FALSE) +
  ggplot2::facet_wrap(~pathway, labeller = ggplot2::label_wrap_gen(40)) +
  ggplot2::labs(x = NULL, y = "Pathway activation") +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 30, hjust = 1))
```

Activations are only defined up to sign and scale, and, like importance scores,
they should be checked for stability across seeds before being interpreted
biologically.

# Survival with clinical covariates

A `Surv()` outcome gives a Cox proportional hazards model, trained by
maximising the partial likelihood. Clinical covariates on the right-hand side
enter at the prediction head alongside the pathway embeddings (late fusion):

```{r surv-fit}
fit_os <- bn_fit(
  Surv(os_time, os_event) ~ age + stage,
  data = brca$clinical[train, ],
  omics = rows(train),
  graph = g,
  control = bn_control(seed = 1)
)
fit_os
```

```{r surv-eval}
te <- brca$clinical[test, ]
ok <- complete.cases(te[c("age", "stage", "os_time")])
lp <- predict(fit_os, rows(test[ok]), covariates = te[ok, ])
concordance(
  Surv(os_time, os_event) ~ lp$.pred_linear_pred,
  data = te[ok, ],
  reverse = TRUE
)$concordance
```

Survival curves come from the Breslow estimate of the baseline hazard:

```{r surv-curves}
s <- predict(
  fit_os,
  rows(test[ok][1:3]),
  covariates = te[ok, ][1:3, ],
  type = "survival",
  eval_time = c(1, 3, 5) * 365
)
s$.pred[[1]]
```

# Regression

A numeric outcome gives regression. For example, how well can copy number
and mutations predict expression of the proliferation marker *TOP2A*?

```{r regression}
y <- brca$omics$rna[, "TOP2A"]
fit_reg <- bn_fit(
  rows(train)[c("cnv", "mut")],
  y[train],
  graph = g,
  control = bn_control(seed = 1)
)
cor(predict(fit_reg, rows(test)[c("cnv", "mut")])$.pred, y[test])
```

# A second cohort: prostate cancer grade

The same knowledge graph serves any cohort; `bn_fit()` prunes it to the genes
that are observed. `tcga_cohort("PRAD")` holds RNA and copy number for 491 prostate tumours
with Gleason grade, the task used to showcase P-NET. A binary factor gives a
two-class model, and `type = "prob"` returns the class probabilities needed for
an ROC curve:

```{r prad}
prad <- tcga_cohort("PRAD")
set.seed(7)
test_p <- sample(nrow(prad$clinical), 150)
train_p <- setdiff(seq_len(nrow(prad$clinical)), test_p)
fit_prad <- bn_fit(
  grade ~ 1,
  data = prad$clinical[train_p, ],
  omics = lapply(prad$omics, function(m) m[train_p, ]),
  graph = g,
  control = bn_control(seed = 1)
)
fit_prad
p_high <- predict(
  fit_prad,
  lapply(prad$omics, function(m) m[test_p, ]),
  type = "prob"
)$.pred_high
truth <- prad$clinical$grade[test_p] == "high"
# AUROC via the Mann-Whitney statistic (or use pROC / yardstick)
(sum(rank(p_high)[truth]) - sum(truth) * (sum(truth) + 1) / 2) /
  (sum(truth) * sum(!truth))
```

# Beyond BINNs: relaxing the architecture

Each constraint of a feedforward BINN is a single argument of `bn_arch()`:

| BINN assumption | `bn_arch()` argument | relaxation |
|---|---|---|
| one scalar per pathway | `hidden_dim = 1` | vector embeddings, e.g. `hidden_dim = 4` |
| a unique weight per edge | `weights = "edge"` | `"shared"`: mean aggregation + per-level transform (GCN) |
| predict from top pathways | `readout = "roots"` | `"pathways"`: pool all pathway nodes |
| stratified schedule, dummy nodes | (not needed) | topological schedule follows the DAG |
| single fusion point | `gene_layer` | gene nodes merge omics layers first |

The `"gnn"` preset combines 4-dimensional embeddings with shared weights:

```{r gnn}
fit_gnn <- bn_fit(
  Surv(os_time, os_event) ~ age + stage,
  data = brca$clinical[train, ],
  omics = rows(train),
  graph = g,
  arch = bn_arch("gnn"),
  control = bn_control(seed = 1)
)
glance(fit_gnn)
```

## Is it the biology or just the sparsity?

To test whether a model benefits from the *specific* prior knowledge, or merely
from being sparse, fit it on a null graph with the same degree sequence
(`bn_rewire()`) and compare. The benchmark vignette does this systematically
with cross-validation.

```{r null}
g_null <- bn_rewire(g, method = "degree", seed = 1)
fit_null <- bn_fit(
  Surv(os_time, os_event) ~ age + stage,
  data = brca$clinical[train, ],
  omics = rows(train),
  graph = g_null,
  control = bn_control(seed = 1)
)
lp_null <- predict(fit_null, rows(test[ok]), covariates = te[ok, ])
concordance(
  Surv(os_time, os_event) ~ lp_null$.pred_linear_pred,
  data = te[ok, ],
  reverse = TRUE
)$concordance
```

## Incomplete knowledge graphs

Pathway databases are incomplete. Because the graph is data, the effect of
missing annotations can be simulated directly: `bn_drop_edges()` removes a
random share of gene-pathway annotations, and the model is refitted.

```{r incomplete}
g_half <- bn_drop_edges(g, prop = 0.5, seed = 1)
summary(g_half)
fit_half <- bn_fit(
  pam50 ~ 1,
  data = brca$clinical[train, ],
  omics = rows(train),
  graph = g_half,
  control = bn_control(seed = 1)
)
mean(
  predict(fit_half, rows(test_ok))$.pred_class == brca$clinical$pam50[test_ok]
)
```

Genes that lose all their annotations drop out of the model entirely, so an
incomplete graph also changes which features are used.

# Working with Bioconductor objects

If your data are in a `MultiAssayExperiment`, pass it as `data`. The omics
layers and clinical variables (`colData`) are extracted and matched by patient
automatically. Assays must be matrix-like with gene symbols as row names
(use `mae_to_omics()` to see exactly what will be used).

```{r mae, eval = run && has_data && requireNamespace("MultiAssayExperiment", quietly = TRUE)}
suppressPackageStartupMessages(library(MultiAssayExperiment))
cd <- as.data.frame(brca$clinical)
rownames(cd) <- cd$patient
mae <- MultiAssayExperiment(
  experiments = list(rna = t(brca$omics$rna), cnv = t(brca$omics$cnv)),
  colData = S4Vectors::DataFrame(cd)
)
mae
fit_mae <- bn_fit(
  pam50 ~ age,
  data = mae,
  graph = g,
  control = bn_control(epochs = 20, seed = 1)
)
fit_mae
```

For TCGA data, `curatedTCGAData::curatedTCGAData()` returns such objects
directly. See `data-raw/tcga.R` in the package sources for how `brca` was
built.

# Current limitations

This is an alpha release. Not yet supported:

* missing values or missing omics layers (drop or impute them first);
* pathway expansion or learning new edges;
* mid-level fusion of clinical data.

The graph formulation makes all of these natural extensions.

# Session information {.unnumbered}

```{r session, eval = TRUE}
sessionInfo()
```
